# Reproducible Research: Peer Assessment 1
Author: Christian J. Lagares Nieves


## Loading and preprocessing the data
The data was obtained from the following link `https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip` on July 13, 2015 nearing 12:00 PM GMT-4. Data was provided solely for the purpose of analysis and it is not attributed to the author of this document or his affiliates. 

```{r loaddata, echo = TRUE}
unzip(zipfile="repdata-data-activity.zip")
data <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?

```{r, echo = TRUE}
library(ggplot2)
total.steps <- tapply(data$steps, data$date, FUN=sum, na.rm=TRUE)
qplot(total.steps, binwidth=1000, xlab="Total Number of Steps Each Day")
mean(total.steps, na.rm=TRUE)
median(total.steps, na.rm=TRUE)
```

## What is the average daily activity pattern?
   Patterns help us find answers to important questions or guide us to a more appropiate course of action towards finding the answer.
```{r, echo = TRUE}
library(ggplot2)
averages <- aggregate(x=list(steps=data$steps), by=list(interval=data$interval),
                      FUN=mean, na.rm=TRUE)
ggplot(data=averages, aes(x=interval, y=steps)) +
    geom_line() +
    xlab("5-minute Interval") +
    ylab("Average Number of Steps Taken")
```

   On average across all the days in the dataset, the 5-minute interval contains
the maximum number of steps?
```{r, echo = TRUE}
averages[which.max(averages$steps),]
```

## Imputing missing values
Missing data should always be taken into account and is a crucial factor with data acquiered through non-constant methodology like personal training meters.

There are many days/intervals where there are missing values (coded as `NA`). The presence of missing days may introduce bias into some calculations or summaries of the data.

```{r how_many_missing, echo = TRUE}
missing <- is.na(data$steps)
table(missing)
```

An acceptable and rather simple approach could be employing the 5-minute interval mean value. In the following segment, all of the missing values are filled in with mean value for that 5-minute interval.

```{r, echo = TRUE}
# This segment replaces each missing value with the mean value of its 5-minute interval
fill.value <- function(steps, interval) {
    filled <- NA
    if (!is.na(steps))
        filled <- c(steps)
    else
        filled <- (averages[averages$interval==interval, "steps"])
    return(filled)
}
filled.data <- data
filled.data$steps <- mapply(fill.value, filled.data$steps, filled.data$interval)
```
Now, using the filled data set, let's make a histogram of the total number of steps taken each day and calculate the mean and median total number of steps. This should help us summarize the data into some actual facts.

```{r, echo = TRUE}
total.steps <- tapply(filled.data$steps, filled.data$date, FUN=sum)
qplot(total.steps, binwidth=1000, xlab="Total Number of Steps Taken Each Day")
mean(total.steps)
median(total.steps)
```

Mean and median values are higher after imputing missing data. The reason is
that in the original data, there are some days with `steps` values `NA` for 
any `interval`. The total number of steps taken in such days are set to 0s by
default. However, after replacing missing `steps` values with the mean `steps`
of associated `interval` value, these 0 values are removed from the histogram
of total number of steps taken each day.

## Are there differences in activity patterns between weekdays and weekends?
First, let's find the day of the week for each measurement in the dataset. In
this part, we use the dataset with the filled-in values.

```{r, echo = TRUE}
weekday.or.weekend <- function(date) {
    day <- weekdays(date)
    # The following selection structure is a simple segment of code that helps us designate a category to the week days.
    if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
        return("weekday")
    else if (day %in% c("Saturday", "Sunday"))
        return("weekend")
    else
        stop("invalid date")
}
filled.data$date <- as.Date(filled.data$date)
filled.data$day <- sapply(filled.data$date, FUN=weekday.or.weekend)
```

Now, let's make a panel plot containing plots of average number of steps taken
on weekdays and weekends.
```{r, echo = TRUE}
averages <- aggregate(steps ~ interval + day, data=filled.data, mean)
ggplot(averages, aes(interval, steps)) + geom_line() + facet_grid(day ~ .) +
    xlab("5-minute interval") + ylab("Number of steps")
```